D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-04-06 06related_conclusion.tex
.

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,coecke2011interacting}
is a graphical calculus for quantum states. A main appeal of the ZX
calculus is that its foundations are grounded in categorical quantum
mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
modeling and reasoning about quantum physics. Another main appeal of
the ZX calculus is that ZX diagrams have an operational interpretation
using graph rewriting. There is a large body of work that studies
properties of these rewriting systems, and in particular completeness
(a set of rules is complete if it can prove all valid identities) and
minimality (a complete set of rules is minimal if removing a rule
leads to incompleteness). Two main proof techniques for completeness
are via termination, or via interpretation. In the first case, one
shows that a rewriting system has unique normal forms and that two
expressions are semantically equivalent iff they have the same normal
form---there exists many relaxations of this result---whereas in the
second case one shows completeness by exhibiting a well-behaved
translation to another system for which completeness holds. Both proof
techniques have been used to prove completeness for multiple settings,
including the Clifford fragment~\cite{backens2014zx}, the
Toffoli+Hadamard fragment~\cite{hadzi15zx}, the Clifford+T
fragment~\cite{jeandel2018complete} and the (qubit) universal
fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
which encompasses the ZX and ZW calculi~\cite{zwcalculus}. We refer
the reader to~\cite{vandewetering2020zx} for a (relatively) recent
historical and technical account of completeness results.

Because they are both grounded in rewriting techniques, there are many
similarities between prior work on the ZX calculus and our own work on
labelled Dirac notation. We highlight some key differences. First,
there exist a number of differences between our language and the
concrete fragments of the ZX-calculus for which completeness has been
studied. For instance, our language features big operators, and
operators for dual spaces. A second difference is that we do not claim
completeness: our main theorem shows a reduction from Labelled Dirac
notation to Dirac notation, for which completeness is only partially
known.

it is difficult to compare the ZX
calculus and LDN. First, they have different scopes: the ZX calculus
mainly focuses on quantum circuits, whereas LDN aims to provide
general support for quantum physics. This difference in scope is
reflected in the underlying languages: for instance, many works in ZX
calculus consider concrete sets of quantum gates, whereas LDN
emphasizes the use of big operators, and does not attempt to provide
specialized support for a particular set of quantum gates. As a
result, many of the examples handled by D-Hammer cannot be formalized
in existing ZX-calculus tools. On the contrary, D-Hammer is not
competitive on the examples that can be verified by ZX-calculus.

\gb{give a few examples in terms of size and time}

\paragraph*{Comparison with path-over sums approach}
Similarly, there exist
completeness results and tools for the sum-over-paths representation
of quantum states~\cite{amy2018towards,amy2023complete}.

\paragraph*{Comparison with equivalence checking}

Automata-based approaches has also been successfully applied for
reasoning about quantum circuits~\cite{AutoQ_pldi_2023}. Instead of
representing a set of $n$-qubit states in exponential size,
\cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much
less (often $\mathit{poly}(n)$) transitions.  They further established
transformers to implement the semantics of quantum gates over this
representation, providing better scalability compared to other
approaches. It is developed as a practical tool \textsc{autoq} with
the capabilities for symbolic reasoning~\cite{AutoQ2023}. More
recently, \cite{AutoQ_popl2025} further explore level-synchronized
tree automata to encoding state, a promising model for parameterized
verification.

\paragraph*{Canonical forms in multi-body quantum physics}

>>>>

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,coecke2011interacting}
is a graphical calculus for quantum states. A main appeal of the ZX
calculus is that its foundations are grounded in categorical quantum
mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
modeling and reasoning about quantum physics. Another main appeal of
the ZX calculus is that ZX diagrams have an operational interpretation
using graph rewriting. There is a large body of work that studies
properties of these rewriting systems, and in particular completeness
(a set of rules is complete if it can prove all valid identities) and
minimality (a complete set of rules is minimal if removing a rule
leads to incompleteness). Two main proof techniques for completeness
are via termination, or via interpretation. In the first case, one
shows that a rewriting system has unique normal forms and that two
expressions are semantically equivalent iff they have the same normal
form---there exists many relaxations of this result---whereas in the
second case one shows completeness by exhibiting a well-behaved
translation to another system for which completeness holds. Both proof
techniques have been used to prove completeness for multiple settings,
including the Clifford fragment~\cite{backens2014zx}, the
Toffoli+Hadamard fragment~\cite{hadzi15zx}, the Clifford+T
fragment~\cite{jeandel2018complete} and the (qubit) universal
fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
which encompasses the ZX and ZW calculi~\cite{zwcalculus}, quantum
circuits~\cite{DBLP:conf/lics/ClementHMPV23} or optical quantum
circuits~\cite{clement_et_al:2022}. We refer the interested reader
to~\cite{vandewetering2020zx} for a historical and technical account
of completeness results up to 2020.

Because they are both grounded in rewriting techniques, there are many
similarities between prior work on the ZX calculus and our own work on
labelled Dirac notation. We highlight some key differences. First,
there exist a number of differences between our language and the
concrete fragments of the ZX-calculus for which completeness has been
studied. For instance, our language features big operators, and
operators for dual spaces. A second difference is that we do not claim
completeness: our main theorem shows a reduction from Labelled Dirac
notation to Dirac notation, for which completeness is only partially
known. From a practical point of view, many of the examples handled by
D-Hammer cannot be formalized in existing ZX-calculus tools. On the
contrary, D-Hammer is not competitive on the examples that can be
verified by ZX-calculus. \gb{give a few examples in terms of size and time}

\paragraph*{Comparison with quantum circuit equivalence checking tools}
A main application of the ZX-calculus is to check equivalence of
quantum circuits; for instance, PyZX~\cite{kissinger2020pyzx} is a
popular Python library for automating equational reasoning in the
ZX-calculus.

There are many other tools for equivalence checking of quantum
circuits. Amy~\cite{amy2018towards} uses the path-sums formalism
formalism to check circuits with thousands of T gates. A series of
works~\cite{AutoQ_pldi_2023,AutoQ2023,AutoQ_popl2025} explores
automata-based approaches to verify quantum circuits at scale. Their
main idea is to encode circuits as a compact, finite tree automata,
often with $\mathit{poly}(n)$ transitions (instead of $2^n$
transitions), where $n$ is the number of qubits of the state. By
providing the semantics of quantum gates over this representation,
they achieve practical verification. Their tool \textsc{autoq} is able
to verify circuits with over 100,00 gates, and was recently extended
to support parametrized verification.

\paragraph*{Canonical forms in multi-body quantum physics}
Canonical forms play a fundamental role in quantum physics. For
instance, \cite{perezgarcia2007,RevModPhys.93.045003,Acuaviva_2023}
discuss canonical forms of Matrix Product States (MPS) and Tensor
Networks respectively. An exciting direction for further work is to
further develop automated deduction techniques for quantum physics.
